---
title: "plyxp usage examples"
author:
    - name: Justin Landis
      orcid: 0000-0001-5501-4934
    - name: Michael Love
      orcid: 0000-0001-8401-0545
date: "`r format(Sys.time(), '%m/%d/%Y')`"
format:
  pdf:
    geometry: margin=1in
    toc: true
    number-sections: true
    colorlinks: true
    include-in-header:
      text: |
        \usepackage{etoolbox}
        \AtBeginEnvironment{verbatim}{\small}
---

# Introduction

This document is a modified version of the *plyxp* vignette 
as a supplement for the *plyxp* paper.

The full and up-to-date *plyxp* vignette can be found here:

<https://jtlandis.github.io/plyxp/articles/plyxp.html>

To install *plyxp* use the following code:

```{r}
#| eval: false
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("plyxp")
```

Information about *plyxp* can be found at the following two locations:

  - <https://github.com/jtlandis/plyxp>
  - <https://bioconductor.org/packages/plyxp>

```{r}
#| include: false
options(width = 80)
```

# plyxp use cases

## Basic manipulation of an RNA-seq dataset

We load the *airway* dataset which contains a *SummarizedExperiment* 
object with gene-level counts across eight samples.
The *airway* dataset summarizes an RNA-Seq experiment on four human airway 
smooth muscle cell lines, both as controls and treated with dexamethasone.
Details on the gene model and read counting procedure are provided in the *airway* 
package vignette from Bioconductor.
The citation for the experiment is: 

> Himes BE, Jiang X, Wagner P, Hu R, Wang Q, Klanderman B, Whitaker RM, 
> Duan Q, Lasky-Su J, Nikolos C, Jester W, Johnson M, Panettieri R Jr, 
> Tantisira KG, Weiss ST, Lu Q. 
> "RNA-Seq Transcriptome Profiling Identifies CRISPLD2 as a Glucocorticoid 
> Responsive Gene that Modulates Cytokine Function in Airway Smooth Muscle Cells."
> *PLOS One*. 2014 Jun 13;9(6):e99625. PMID: 24926665. GEO: GSE52778.

```{r}
#| message: false
library(airway)
data(airway)
library(dplyr)
```

```{r}
# cleaning up the metadata...
colData(airway) <- colData(airway)[,1:4]
rowData(airway) <- rowData(airway)[,c(1:2,4)]
rowRanges(airway) <- NULL # ranges not needed
```

```{r}
#| message: false
library(plyxp)
xp <- new_plyxp(airway)
```

```{r}
#| size: small
xp |>
  mutate(log_counts = log1p(counts),
         cols(treated = dex == "trt"),
         rows(new_id = paste0("gene-", gene_name)))
```

```{r}
xp$sizeFactor <- runif(8, .9, 1.1)
xp |>
  mutate(scaled_counts = counts / .cols$sizeFactor, #
         rows(ave_scl_cts = rowMeans(.assays_asis$scaled_counts)))
```

```{r}
summary <- xp |>
  group_by(rows(gene_biotype)) |>
  summarize(col_sums = colSums(counts),
            # may rename rows with .features
            rows(.features = unique(gene_biotype)))
summary |> rowData()
```

```{r}
library(tibble)
summary |>
  pull(col_sums) |>
  as_tibble(rownames = "type")
```

To extract the *SummarizedExperiment*:

```{r}
se(xp)
```

## Exploring data alongside DE results

Mike will add something here...

# Session Info

```{r}
sessionInfo()
```
