---
title: "plyxp usage examples"
author:
  - name: Justin Landis
    orcid: 0000-0001-5501-4934
  - name: Michael Love
    orcid: 0000-0001-8401-0545
date: "`r format(Sys.time(), '%m/%d/%Y')`"
format:
  pdf:
    geometry: margin=1in
    toc: true
    number-sections: true
    colorlinks: true
    include-in-header:
      text: |
        \usepackage{etoolbox}
        \AtBeginEnvironment{verbatim}{\small}
---

# Introduction

This document is a modified version of the _plyxp_ vignette
as a supplement for the _plyxp_ paper.

The full and up-to-date _plyxp_ vignette can be found here:

<https://jtlandis.github.io/plyxp/articles/plyxp.html>

To install _plyxp_ use the following code:

```{r}
#| eval: false
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("plyxp")
```

Information about _plyxp_ can be found at the following two locations:

- <https://github.com/jtlandis/plyxp>
- <https://bioconductor.org/packages/plyxp>

```{r}
#| include: false
options(width = 80)
```

# Load an RNA-seq data object

We load the _airway_ dataset which contains a _SummarizedExperiment_
object with gene-level counts across eight samples.
The _airway_ dataset summarizes an RNA-Seq experiment on four human airway
smooth muscle cell lines, both as controls and treated with dexamethasone.
Details on the gene model and read counting procedure are provided in the _airway_
package vignette from Bioconductor. For more details on construction of
_SummarizedExperiment_ (SE) objects, see the vignette for the package
of the same name, or the _tximeta_ package which can construct these objects
from common RNA-seq quantification tools.

The citation for the experiment is:

> Himes BE, Jiang X, Wagner P, Hu R, Wang Q, Klanderman B, Whitaker RM,
> Duan Q, Lasky-Su J, Nikolos C, Jester W, Johnson M, Panettieri R Jr,
> Tantisira KG, Weiss ST, Lu Q.
> "RNA-Seq Transcriptome Profiling Identifies CRISPLD2 as a Glucocorticoid
> Responsive Gene that Modulates Cytokine Function in Airway Smooth Muscle Cells."
> _PLOS One_. 2014 Jun 13;9(6):e99625. PMID: 24926665. GEO: GSE52778.

```{r}
#| message: false
library(airway)
data(airway)
```

First we prepare the `airway` object a bit, by removing extra variables
from the row and column metadata.

```{r}
# cleaning up the metadata...
colData(airway) <- colData(airway)[,c("cell","dex")]
rowData(airway) <- rowData(airway)[,c("gene_id","symbol","gene_biotype")]
rowData(airway)$gene_length <- sum(width(reduce(rowRanges(airway))))
rowRanges(airway) <- NULL # ranges not needed for this demo, simplifying...
```

# Basic manipulation of RNA-seq data

We initialize by calling `new_plyxp()` on an SE object.
This is a simple wrapper class that contains the SE, and
allows for us to change the behavior of the dplyr verbs
when they are applied to our new object.

```{r}
#| message: false
library(plyxp)
xp <- new_plyxp(airway)
class(xp)
```

We can perform one or more operations, modifying the
contents of the object:

```{r}
xp |>
  mutate(
    log10_cts = log10(counts + 1),
    cols(treated = dex == "trt"),
    rows(gene_kb = gene_length / 1e3)
  )
```

The above computations only made use of data
_within_ each context. Using pronouns, we can also
access data _across_ contexts:

```{r}
xp |>
  mutate(
    cols(col_sum_per_mil = colSums(.assays_asis$counts)/1e6),
    scaled_counts = counts/.cols$col_sum_per_mil,
    rows(ave_scl_cts = rowMeans(.assays_asis$scaled_counts))
  )
```

We can also perform grouping operations, e.g. here grouping by the
gene biotype, and computing the sum of counts for each sample:

```{r}
summary <- xp |>
  group_by(rows(gene_biotype)) |>
  summarize(
    col_sums = colSums(counts),
    # may rename rows with .features
    rows(.features = unique(gene_biotype))
  )
```

Filtering to a single sample, and examining the first six biotypes:

```{r}
summary |>
  filter(cols(.samples == "SRR1039508")) |>
  head()
```

We can convert the SE to tibble after selecting one of the assays:

```{r}
library(tibble)
summary |>
  pull(col_sums) |>
  as_tibble(rownames = "type") |>
  head()
```

To extract the _SummarizedExperiment_ from a
_PlySummarizedExperiment_ we just use the `se()` function:

```{r}
se(xp)
```

# Exploring data alongside DE results

```{r}
# define minimal expression as count >= 10
filt_xp <- xp |>
  mutate(rows(
    min_exprs_10 = rowSums(.assays_asis$counts >= 10)
  )) |>
  filter(rows(min_exprs_10 >= 4)) |>
  mutate(cols(
    dex = factor(dex, levels=c("untrt","trt"))
  ))
```

```{r}
#| message: false
library(DESeq2)
dds <- se(filt_xp) |>
  DESeqDataSet(design=~cell + dex) |>
  DESeq(test="LRT", reduced=~cell, quiet=TRUE)
```

```{r}
cols_to_save <- c(
  "gene_id","symbol","gene_biotype","gene_length",
  "baseMean","dispersion","dex_trt_vs_untrt","LRTPvalue"
)
rowData(dds) <- rowData(dds)[,cols_to_save]
assays(dds) <- assays(dds)[c("counts","mu")]
```

```{r}
de_xp <- dds |> new_plyxp()
de_xp <- de_xp |>
  mutate(rows(
    padj = p.adjust(LRTPvalue, method="BH"),
    lfc_sign = sign(dex_trt_vs_untrt)
  ))
rowData(de_xp)
```

```{r}
with(rowData(de_xp),
  table(lfc_sign, sig_5pct = padj < .05)
)
```

# Computing and plotting abundance

```{r}
de_tpm <- de_xp |>
  mutate(
    RPK = counts / (.rows$gene_length / 1e3),
    cols(rpk_scale = colSums(.assays_asis$RPK) / 1e6),
    TPM = RPK / .cols$rpk_scale
  )
de_tpm
```

We can directly pipe an SE object into a `ggplot()` command
after loading _tidySummarizedExperiment_.

```{r}
#| message: false
library(tidySummarizedExperiment)
library(ggplot2)

biotypes <- c("protein_coding", "lincRNA", "antisense")
de_tpm |>
  filter(rows(gene_biotype %in% biotypes)) |>
  se() |>
  ggplot(aes(x = TPM + 0.1, y = gene_biotype)) +
  geom_violin(aes(fill = dex), color = NA) +
  scale_x_log10(labels = scales::label_log())
```

# Grouping and summarizing abundance over samples

First we filter to significant genes and group samples by
condition group, the summarize the average TPM in each group.

```{r}
de_tpm_summary <- de_tpm |>
  filter(rows(padj < 0.05)) |>
  group_by(cols(dex)) |>
  summarize(ave_TPM = rowMeans(TPM), cols(.samples = unique(dex)))

de_tpm_summary
```

Next we group the data by gene biotype and select the top three
genes per biotype by the largest `TPM_ave` difference
between treatment and control groups.

This can be done in two ways, since we know there is only two treatment groups,
we may use `purrr::map_dbl` on the reshaped variant of `ave_TPM`, which returns
a list of row slices, or in other words, a list of $1 \times 2$ matrices.

```{r}

de_tpm_summary |>
  group_by(rows(gene_biotype)) |>
  arrange(rows(
    purrr::map_dbl(
      # list of 1x2 matrices
      .assays$ave_TPM,
      # abs(.x[2] - .x[1])
      ~ abs(diff(.x))
    ) |>
      dplyr::desc()
  )) |>
  slice(rows(1:3))
```

The above may not always be preferable without first knowing the order of
the summarized columns. The summarized dimensions are ordered based on the
appearance of the values from the original data.
Alternatively, if you had more than 2 treatment groups and/or needed to be more
specific on which columns were used in the difference, then you can use
the `.assays_asis` variant as shown below.

```{r}
de_tpm_summary_slice <- de_tpm_summary |>
  group_by(rows(gene_biotype)) |>
  arrange(rows(
    {
      mat <- .assays_asis$ave_TPM
      abs(mat[,"trt"] - mat[,"untrt"])
    } |>
     dplyr::desc()
  )) |>
  slice(rows(1:3))

de_tpm_summary_slice
```

Finally, we can make a plot of top 3 TPM differences per biotype, with
a point for the median and a line representing the range within the group:

```{r}
de_tpm |>
  filter(rows(gene_biotype %in% biotypes)) |>
  filter(rows(gene_id %in% rownames(de_tpm_summary_slice))) |>
  se() |>
  ggplot(aes(x = TPM + 0.1, y = symbol, group = dex, color = dex)) +
  stat_summary(
    fun.min = min, fun.max = max, fun = mean,
    geom = "pointrange", position = position_dodge(width = 1)
  ) +
  facet_grid(rows = vars(gene_biotype), scales = "free_y") +
  scale_x_log10(labels = scales::label_log())
```

We can also make a simple heatmap of similar data:

```{r}
de_tpm |>
  filter(rows(gene_biotype %in% biotypes)) |>
  filter(rows(gene_id %in% rownames(de_tpm_summary_slice))) |>
  se() |>
  ggplot(aes(x = dex, y = symbol)) +
  geom_tile(aes(fill = log10(TPM + 0.1))) +
  facet_grid(rows = vars(gene_biotype), cols = vars(cell), scales = "free_y") +
  scale_fill_viridis_c()
```

# Session Info

```{r}
sessionInfo()
```
