---
title: "plyxp usage examples"
author:
  - name: Justin Landis
    orcid: 0000-0001-5501-4934
  - name: Michael Love
    orcid: 0000-0001-8401-0545
date: "`r format(Sys.time(), '%m/%d/%Y')`"
format:
  pdf:
    geometry: margin=1in
    toc: true
    number-sections: true
    colorlinks: true
    include-in-header:
      text: |
        \usepackage{etoolbox}
        \AtBeginEnvironment{verbatim}{\small}
---

# Introduction

This document is a modified version of the _plyxp_ vignette
as a supplement for the _plyxp_ paper.

The full and up-to-date _plyxp_ vignette can be found here:

<https://jtlandis.github.io/plyxp/articles/plyxp.html>

To install _plyxp_ use the following code:

```{r}
#| eval: false
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("plyxp")
```

Information about _plyxp_ can be found at the following two locations:

- <https://github.com/jtlandis/plyxp>
- <https://bioconductor.org/packages/plyxp>

```{r}
#| include: false
options(width = 80)
```

# plyxp use cases

## Basic manipulation of an RNA-seq dataset

We load the _airway_ dataset which contains a _SummarizedExperiment_
object with gene-level counts across eight samples.
The _airway_ dataset summarizes an RNA-Seq experiment on four human airway
smooth muscle cell lines, both as controls and treated with dexamethasone.
Details on the gene model and read counting procedure are provided in the _airway_
package vignette from Bioconductor.
The citation for the experiment is:

> Himes BE, Jiang X, Wagner P, Hu R, Wang Q, Klanderman B, Whitaker RM,
> Duan Q, Lasky-Su J, Nikolos C, Jester W, Johnson M, Panettieri R Jr,
> Tantisira KG, Weiss ST, Lu Q.
> "RNA-Seq Transcriptome Profiling Identifies CRISPLD2 as a Glucocorticoid
> Responsive Gene that Modulates Cytokine Function in Airway Smooth Muscle Cells."
> _PLOS One_. 2014 Jun 13;9(6):e99625. PMID: 24926665. GEO: GSE52778.

```{r}
#| message: false
library(airway)
data(airway)
```

First we prepare the `airway` object a bit, by removing extra variables
from the row and column metadata.

```{r}
# cleaning up the metadata...
colData(airway) <- colData(airway)[,c("cell","dex")]
rowData(airway) <- rowData(airway)[,c("gene_id","symbol","gene_biotype")]
rowData(airway)$gene_length <- sum(width(reduce(rowRanges(airway))))
rowRanges(airway) <- NULL # ranges not needed for this demo, simplifying...
```

```{r}
#| message: false
library(plyxp)
xp <- new_plyxp(airway)
```

```{r}
xp |>
  mutate(
    log10_cts = log10(counts + 1),
    cols(treated = dex == "trt"),
    rows(gene_kb = gene_length / 1e3)
  )
```

```{r}
xp |>
  mutate(
    cols(col_sum_per_mil = colSums(.assays_asis$counts)/1e6),
    scaled_counts = counts/.cols$col_sum_per_mil,
    rows(ave_scl_cts = rowMeans(.assays_asis$scaled_counts))
  )
```

```{r}
summary <- xp |>
  group_by(rows(gene_biotype)) |>
  summarize(
    col_sums = colSums(counts),
    # may rename rows with .features
    rows(.features = unique(gene_biotype))
  )
summary |> rowData()
```

```{r}
library(tibble)
summary |>
  pull(col_sums) |>
  as_tibble(rownames = "type")
```

To extract the _SummarizedExperiment_:

```{r}
se(xp)
```

## Exploring data alongside DE results

```{r}
# define minimal expression as count >= 10
filt_xp <- xp |>
  mutate(rows(
    min_exprs_10 = rowSums(.assays_asis$counts >= 10)
  )) |>
  filter(rows(min_exprs_10 >= 4)) |>
  mutate(cols(
    dex = factor(dex, levels=c("untrt","trt"))
  ))
```

```{r}
#| message: false
library(DESeq2)
dds <- se(filt_xp) |>
  DESeqDataSet(~cell + dex) |>
  DESeq(test="LRT", reduced=~cell, quiet=TRUE)
```

```{r}
cols_to_save <- c(
  "gene_id","symbol","gene_biotype","gene_length",
  "baseMean","dispersion","dex_trt_vs_untrt","LRTPvalue"
)
rowData(dds) <- rowData(dds)[,cols_to_save]
assays(dds) <- assays(dds)[c("counts","mu")]
```

```{r}
de_xp <- dds |> new_plyxp()
de_xp <- de_xp |>
  mutate(rows(
    padj = p.adjust(LRTPvalue, method="BH"),
    lfc_sign = sign(dex_trt_vs_untrt)
  ))
rowData(de_xp)
```

```{r}
with(rowData(de_xp),
  table(lfc_sign, sig_5pct = padj < .05)
)
```

# Compute Log10 TPM

```{r}
de_tpm <- de_xp |>
  mutate(
    RPK = counts / (.rows$gene_length / 1e3),
    cols(rpk_scale = colSums(.assays_asis$RPK) / 1e6),
    TPM = RPK / .cols$rpk_scale
  )
de_tpm
```

```{r}
library(tidySummarizedExperiment)
library(ggplot2)

se(de_tpm) |>
  ggplot(aes(x = TPM, y = gene_biotype)) +
  scale_x_log10()
```

```{r}
# significant genes with maximal TPM_ave diff
# between control and treatment conditions
de_tpm_summary <- de_tpm |>
  filter(rows(padj < 0.05)) |>
  group_by(cols(dex)) |>
  summarize(TPM_ave = rowMeans(TPM), cols(.samples = unique(dex))) |>
  group_by(rows(gene_biotype)) |>
  arrange(rows(- purrr::map_dbl(.assays$TPM_ave, ~ abs(diff(.x))))) |>
  slice(rows(1:3))
de_tpm_summary
```

```{r}

# voilin plot of top 3 TPM_ave differences
de_tpm |>
  filter(rows(gene_id %in% rownames(de_tpm_summary))) |>
  se() |>
  ggplot(aes(x = TPM, y = gene_id)) +
  geom_violin(aes(fill = dex), color = NA) +
  facet_grid(rows = vars(gene_biotype), scales = "free_y") +
  geom_point(
    data = as_tibble(de_tpm_summary),
    aes(x = TPM_ave, group = dex), position = position_dodge(width = 1)
  ) +
  scale_x_log10()

# simple heatmap of similar data
de_tpm |>
  filter(rows(gene_id %in% rownames(de_tpm_summary))) |>
  se() |>
  ggplot(aes(x = dex, y = gene_id)) +
  geom_tile(aes(fill = log10(TPM))) +
  facet_grid(rows = vars(gene_biotype), scales = "free_y") +
  scale_fill_viridis_c()

```

# Session Info

```{r}
sessionInfo()
```
