---
title: "plyxp usage vignette"
author:
  - name: Justin Landis
    orcid: 0000-0001-5501-4934
  - name: Michael Love
    orcid: 0000-0001-8401-0545
date: "`r format(Sys.time(), '%m/%d/%Y')`"
format:
  pdf:
    geometry: margin=1in
    toc: true
    number-sections: true
    colorlinks: true
    include-in-header:
      text: |
        \usepackage{etoolbox}
        \AtBeginEnvironment{verbatim}{\small}
---

# Introduction

This document is a modified version of the _plyxp_ vignette
as a supplement for the _plyxp_ paper. The `.qmd` file that
generated this document can be found here:

<https://github.com/jtlandis/plyxp_paper>

The full and up-to-date _plyxp_ vignette can be found here:

<https://jtlandis.github.io/plyxp/articles/plyxp.html>

To install _plyxp_ use the following code:

```{r}
#| eval: false
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("plyxp")
```

Information about _plyxp_ can be found at the following two locations:

- <https://github.com/jtlandis/plyxp>
- <https://bioconductor.org/packages/plyxp>

```{r}
#| include: false
options(width = 80)
```

# Load an RNA-seq data object

We load the _airway_ dataset which contains a _SummarizedExperiment_
object with gene-level counts across eight samples.
The _airway_ dataset summarizes an RNA-Seq experiment on four human airway
smooth muscle cell lines, both as controls and treated with dexamethasone.
Details on the gene model and read counting procedure are provided in the _airway_
Bioconductor package vignette. For more details on construction of
_SummarizedExperiment_ (SE) objects, see the vignette for the package
of the same name, or the _tximeta_ package which can construct these objects
from output of common RNA-seq quantification tools.

The citation for the experiment is:

> Himes BE, Jiang X, Wagner P, Hu R, Wang Q, Klanderman B, Whitaker RM,
> Duan Q, Lasky-Su J, Nikolos C, Jester W, Johnson M, Panettieri R Jr,
> Tantisira KG, Weiss ST, Lu Q.
> "RNA-Seq Transcriptome Profiling Identifies CRISPLD2 as a Glucocorticoid
> Responsive Gene that Modulates Cytokine Function in Airway Smooth Muscle Cells."
> _PLOS One_. 2014 Jun 13;9(6):e99625. PMID: 24926665. GEO: GSE52778.

```{r}
#| message: false
library(airway)
data(airway)
```

First we prepare the `airway` object a bit, by removing extra variables
from the row and column metadata.

```{r}
# cleaning up the metadata...
colData(airway) <- colData(airway)[,c("cell","dex")]
rowData(airway) <- rowData(airway)[,c("gene_id","symbol","gene_biotype")]
rowData(airway)$gene_length <- sum(width(reduce(rowRanges(airway))))
rowRanges(airway) <- NULL # ranges not needed for this demo, simplifying...
```

# Basic manipulation of RNA-seq data

We initialize the dataset for use with _plyxp_ by calling `new_plyxp()` on an SE object.
This creates a _PlySummarizedExperiment_, a simple wrapper class that contains the SE,
and changes the behavior of the dplyr verbs when they are applied to our new object.

```{r}
#| message: false
library(plyxp)
xp <- new_plyxp(airway)
class(xp)
```

We can perform one or more operations, modifying the
contents of the object:

```{r}
xp |>
  mutate(
    log10_cts = log10(counts + 1),
    cols(treated = dex == "trt"),
    rows(gene_kb = gene_length / 1e3)
  )
```

The above computations only made use of data
_within_ each context. Using pronouns, we can also
access data _across_ contexts:

```{r}
xp |>
  mutate(
    cols(col_sum_per_mil = colSums(.assays_asis$counts)/1e6),
    scaled_counts = counts/.cols$col_sum_per_mil,
    rows(ave_scl_cts = rowMeans(.assays_asis$scaled_counts))
  )
```

We can also perform grouping operations, e.g. here grouping by the
gene biotype, and computing the sum of counts for each sample:

```{r}
summary <- xp |>
  group_by(rows(gene_biotype)) |>
  summarize(
    col_sums = colSums(counts),
    # may rename rows with .features
    rows(.features = unique(gene_biotype))
  )
```

Filtering to a single sample, and examining the first six biotypes:

```{r}
summary |>
  filter(cols(.samples == "SRR1039508")) |>
  head()
```

We can convert the SE to tibble after selecting one of the assays
(in this case there is only one assay):

```{r}
library(tibble)
summary |>
  pull(col_sums) |>
  as_tibble(rownames = "type") |>
  head()
```

To extract the _SummarizedExperiment_ from a
_PlySummarizedExperiment_ we just use the `se()` function:

```{r}
se(xp)
```

# Exploring data alongside DE results

Below we demonstrate running differential expression (DE)
analysis on the SE object, and then manipulating the SE
alongside results including p-values and log2 fold changes (LFC).

```{r}
# define minimal expression as count >= 10
filt_xp <- xp |>
  mutate(rows(
    min_exprs_10 = rowSums(.assays_asis$counts >= 10)
  )) |>
  filter(rows(min_exprs_10 >= 4)) |>
  mutate(cols(
    dex = factor(dex, levels=c("untrt","trt"))
  ))
```

The following code passes the SE object to the _DESeq2_
functions, creating a _DESeqDataSet_ and then running
a likelihood ratio test. This is standard R/Bioconductor
code, showing that _plyxp_ can easily be used alongside
other Bioconductor packages.

```{r}
#| message: false
library(DESeq2)
dds <- se(filt_xp) |>
  DESeqDataSet(design=~cell + dex) |>
  DESeq(test="LRT", reduced=~cell, quiet=TRUE)
```

We pick a subset of the columns of results in the `rowData`
to use in further analysis:

```{r}
cols_to_save <- c(
  "gene_id","symbol","gene_biotype","gene_length",
  "baseMean","dispersion","dex_trt_vs_untrt","LRTPvalue"
)
rowData(dds) <- rowData(dds)[,cols_to_save]
assays(dds) <- assays(dds)[c("counts","mu")]
```

We can easily pass back into _plyxp_ by wrapping
the _DESeqDataSet_ as a _PlySummarizedExperiment_.
This strategy works for any classes that descend
from SE (e.g. _DESeqDataSet_, _SingleCellExperiment_, etc.).

We then compute adjusted p-values and the sign of the LFC.

```{r}
de_xp <- dds |> new_plyxp()
de_xp <- de_xp |>
  mutate(rows(
    padj = p.adjust(LRTPvalue, method="BH"),
    lfc_sign = sign(dex_trt_vs_untrt)
  ))
rowData(de_xp) |> head()
```

Getting a sense of the number of genes at FDR < 5%:

```{r}
with(rowData(de_xp),
  table(lfc_sign, sig_5pct = padj < .05)
)
```

# Computing and plotting abundance

In the following section, we compute abundance as
_transcripts per million_ (TPM). This involves scaling the
observed counts by the gene length
(here we have access to the number of exonic base pairs),
and then dividing by the column sum to account for sequencing depth
differences. The resulting matrix is multiplied by 1 million,
such that the columns sum to 1 million.

```{r}
de_tpm <- de_xp |>
  mutate(
    RPK = counts / (.rows$gene_length / 1e3),
    cols(rpk_scale = colSums(.assays_asis$RPK) / 1e6),
    TPM = RPK / .cols$rpk_scale
  )

colSums(assay(de_tpm, "TPM"))
```

We can directly pipe an SE object into a `ggplot()` command
after loading _tidySummarizedExperiment_.
We can use this to plot our newly computed TPM values
across genes and samples, splitting by gene biotype,
and by the treatment group.

```{r}
#| message: false
#| warning: false
library(tidySummarizedExperiment)
library(ggplot2)

biotypes <- c("protein_coding", "lincRNA", "antisense")
de_tpm |>
  filter(rows(gene_biotype %in% biotypes)) |>
  se() |>
  ggplot(aes(x = TPM + 0.1, y = gene_biotype)) +
  geom_violin(aes(fill = dex), color = NA) +
  scale_x_log10(labels = scales::label_log())
```

# Grouping and summarizing abundance over samples

Suppose we want to make some plots of the abundance (TPM)
in each treatment group, focusing on the genes
with evidence of differential expression.
First we filter to significant genes and group samples by
condition group, then we can summarize the average TPM in each group,
with the following code:

```{r}

de_tpm_summary <- de_tpm |>
  filter(rows(padj < 0.05)) |>
  group_by(cols(dex)) |>
  summarize(ave_TPM = rowMeans(TPM), cols(.samples = unique(dex)))

de_tpm_summary
```

Next we group the data by gene biotype and select the top three
genes according to the largest `ave_TPM` difference between
treatment and control samples. This gives us back three genes
for each gene biotype.

This can be done in two ways: since we know there are only two treatment groups,
we may use `purrr::map_dbl` on the reshaped variant of `ave_TPM`, which returns
a list of the rows of the `ave_TPM` matrix. In other words, the reshaped data
will be a list of vectors of length 2 (the average TPM for samples in each
treatment group).

```{r}
de_tpm_summary |>
  group_by(rows(gene_biotype)) |>
  arrange(rows(
    purrr::map_dbl(
      # list of length 2
      # numeric vectors
      .assays$ave_TPM,
      ~ abs(diff(.x))
    ) |>
      dplyr::desc()
  )) |>
  slice(rows(1:3))
```

The above approach may not always be ideal if the order of the
summarized columns is not known in advance.
However, because we had renamed the column dimension with
`cols(.samples = unique(dex))`,
we may use the column names as a specific index.
For example, you could use the `.assays_asis` variant as shown below.

```{r}
de_tpm_summary_slice <- de_tpm_summary |>
  group_by(rows(gene_biotype)) |>
  arrange(rows(
    {
      mat <- .assays_asis$ave_TPM
      abs(mat[, "trt"] - mat[, "untrt"])
    } |>
     dplyr::desc()
  )) |>
  slice(rows(1:3))

de_tpm_summary_slice
```

For completeness, we note that this difference could also be computed on the reshaped data,
`map_dbl(.assays$ave_TPM, ~ .x["trt"] - .x["untrt"])`,
though the matrix subtraction above is expected to be faster.

We can make a plot of top 3 TPM differences per biotype, with
a point for the mean and a line representing the range within the group:

```{r}
de_tpm |>
  filter(rows(gene_biotype %in% biotypes)) |>
  filter(rows(gene_id %in% rownames(de_tpm_summary_slice))) |>
  se() |>
  ggplot(aes(x = TPM + 0.1, y = symbol, group = dex, color = dex)) +
  stat_summary(
    fun.min = min, fun.max = max, fun = mean,
    geom = "pointrange", position = position_dodge(width = 1)
  ) +
  facet_grid(rows = vars(gene_biotype), scales = "free_y") +
  scale_x_log10(labels = scales::label_log())
```

We can make a simple heatmap, additionally showing the value per
donor cell line.

```{r}
de_tpm |>
  filter(rows(gene_biotype %in% biotypes)) |>
  filter(rows(gene_id %in% rownames(de_tpm_summary_slice))) |>
  se() |>
  ggplot(aes(x = dex, y = symbol)) +
  geom_tile(aes(fill = log10(TPM + 0.1))) +
  facet_grid(rows = vars(gene_biotype), cols = vars(cell), scales = "free_y") +
  scale_fill_viridis_c()
```

# Session Info

```{r}
sessionInfo()
```
